---
title: "Handout Factors with Forcats"
author: "Nimai Mehta"
date: "`r Sys.Date()`"
output: word_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo       = TRUE, 
                      fig.height = 3, 
                      fig.width  = 6,
                      fig.align  = "center")
ggplot2::theme_set(ggplot2::theme_bw())
```

# Learning Objectives

- Manipulating factors.
- [Wrangling Categorical Data in R](https://doi.org/10.7287/peerj.preprints.3163v2).
- [8.2: Chimeras of the R Inferno](https://www.burns-stat.com/pages/Tutor/R_inferno.pdf)
- [Factors with forcats Cheat Sheet](https://github.com/rstudio/cheatsheets/blob/master/factors.pdf).
- [Forcats Overview](https://forcats.tidyverse.org/).

# Factors

- Factors are used to represent categorical data. Factors can be ordered or unordered and are an important class for statistical analysis and for plotting.

    ```{r}
    suppressPackageStartupMessages(library(tidyverse))
    
    ```
```{r}
sex <- factor(c("male", "female", "female", "male"))
sex
```

```{r}
levels(sex)
nlevels(sex)
```

```{r}
x <- c("51", "32", "15", "2", "32")

xf <- factor(x)
xf
#output shows you levels and position values in environment

as.numeric(xf)

as.numeric("Hello")
as.numeric(factor("Hello"))

##If sure that all levels are numeric and are incorrectly specified as factors, then do this to convert to numeric:

parse_number(levels(xf)[xf]) #[xf] parses it as a numberical output (otherwise just gives you a numerical output)
```

# Creating Factors [Be careful, factors are saved in numerican form]

- Use `factor()` or `parse_factor()` to create a factor variable 

- `parse_factor()`  is similar to `factor()` but will generate warnings if elements of `x` are not found in levels.

    ```{r}
  food <- factor(c("low", "high", "medium", "high", "low", "medium", "high"))
    levels(food)
    ```
    
- If you do not specify the `levels` argument, R will assume that the levels
  are the unique values of the vector.
    - `factor()` takes the order of the levels to be the same order returned 
      by `sort()`.
    - `parse_factor()` takes the order of the levels to be the same order
      as the order of the value introduced.
  
  
    ```{r}
  food <- factor(c("low", "high", "medium", "high", "low", "medium", "high"), levels = c("low","medium","high"), ordered = TRUE)
levels(food)
    ```
```{r}
x <- c("A", "string", "vector", "is", "a", "string", "vector")
factor(x)

sort(unique(x))

parse_factor(x) #gives order of levels in the vector
```

- You can always see the levels of a factor (and their order) using 
  the `levels()` function
  
    ```{r}
    
    ```

- Other options are the `fct_unique()` and `fct_count()` functions from the 
  forcats package.
  
    ```{r}
   fct_unique(food) #ordered list
  
  fct_count(food) #frequency
    ```
    
- You can count the number of levels with `nlevels()`.

    ```{r}
    nlevels(food)
    ```

# forcats

- forcats is an R package which makes two things much easier in R:
    - Changing the order of the levels of the factor variable.
    - Changing the levels of the factor variable.

- It also a few other helper functions for factors.

- All forcat functions begin with `fct_`. So you can type "fct_" then use 
  tab-completion to scroll through the possible functions.
  
- forcats is a part of the tidyverse, so you don't need to load it separately
  when you load the tidyverse.
    

## Changing the Order of the Levels

- Consider the subset of the 
  [General Social Survey](https://en.wikipedia.org/wiki/General_Social_Survey)
  stored in the `gss_cat` data in forcats.
  
    ```{r}
  data("gss_cat") 
  
    ```

- You often want to change the order of the levels of a factor to make 
  plots more insightful.
  
    ```{r}
 tvdata <- gss_cat %>% 
      group_by(relig) %>% 
      summarize(mean_tvhours = mean(tvhours, na.rm = TRUE))

  ggplot(tvdata, aes(x = mean_tvhours, y = relig)) +
    geom_point() +
    xlab("Average TV Hours Watched") +
    ylab("Religion") +
    theme_bw()
  
    
    ```
  
- `fct_reorder()` reorders the levels of a factor according to some values
  of another variable. The arguments are:
    - `f`: The factor vector.
    - `x`: A numeric vector used to reorder the levels.
    - `fun`: A function applied to `x`, the result of which will be used to
      order the levels of `f`.
    
    ```{r}
   levels(tvdata$relig)
    
    ```

- The plot now reorders the y-axis according to the new level order.

    ```{r}
tvdata <- tvdata %>% 
      mutate(relig = fct_reorder(relig, mean_tvhours)) 
    #changes level order based on tv hours
    levels(tvdata$relig)
    
ggplot(tvdata, aes(x = mean_tvhours, y = relig)) +
  geom_point() +
  xlab("Average TV Hours") +
  ylab("Average TV Hours Watched") +
  theme_bw()
    ```

- `fct_rev()` reverses the order of the factors.

    ```{r}
tvdata <- tvdata %>% 
      mutate(relig = fct_rev(relig)) 
    #reverses previous order of levels
    levels(tvdata$relig)
    
ggplot(tvdata, aes(x = mean_tvhours, y = relig)) +
  geom_point() +
  xlab("Average TV Hours") +
  ylab("Average TV Hours Watched") +
  theme_bw()
    
    ```
    
- `fct_relevel()` allows you to move existing levels to any location.

    ```{r}
  levels(tvdata$relig) 
    
    fct_relevel(tvdata$relig, "None") %>% 
      levels()
  #default, brings None to the top
    
     fct_relevel(tvdata$relig, "None", after = nlevels(tvdata$relig)) %>% 
      levels()
     #brings None to the last level (after = [numerical level])
    ```

- **Exercise**: Reorder the levels of the `partyid` variable so that the 
  levels are in alphabetical order.
  
```{r echo=TRUE}
    
levels(gss_cat$partyid)

gss_cat %>% 
      mutate(partyid = fct_relevel(partyid, sort(levels(partyid)))) %>% 
      count(partyid)
    levels(gss_cat$partyid)
    ```


## Modify Factor Levels


- Let's look at the levels of `partyid` in `gss_cat`.

    ```{r}
levels(gss_cat$partyid)
  
    ```
    
- Use `fct_recode()` to change the levels.

    ```{r}
 gss_cat <- gss_cat %>% 
      mutate(partyid = fct_recode(partyid, 
                                  "Republican, Strong" = "Strong Republican",
                                  "Republican, Weak" = "Not str republican",
                                  "Independent, near rep" = "Ind,near rep",
                                  "Independent, near dem" = "Ind,near dem",
                                  "Democrat, Weak" = "Not str democrat",
                                  "Democrat, Strong" = "Strong democrat"))
 levels(gss_cat$partyid)
    ```
  
- New level goes on the left of the equals sign. Old level goes on the right.
  (Just like `mutate()`!)

- **Exercise**: Modify the factor levels of `marital` to be abbreviations of 
  their long-names. For example, "Divorced" can just be "D"

```{r echo=TRUE}
  levels(gss_cat$marital)

gss_cat <- gss_cat %>% 
  mutate(marital = fct_recode(marital, 
                              "NA" = "No answer",
                              "NM" = "Never married",
                              "S" = "Separated",
                              "D" = "Divorced", 
                              "W" = "Widowed", 
                              "M" = "Married"))
    ```

  
## Other Useful Functions.

- `fct_c()`: is the safe way to combine factor vectors.

    ```{r}
   fc1 <- parse_factor(c("A", "B"))
   fc2 <- parse_factor(c("C", "D"))
   
   fct_c(fc1, fc2)
    
    ```

- `fct_collapse()`: combine multiple levels into one level.

    ```{r}
   fc <- parse_factor(c("A", "B", "C", "A", "B", "C"))
    
  fct_collapse(fc, "blah" = c("A", "B"))
    
    ```

- `fct_drop()`: removes any levels that are unused.
  
    ```{r}
   fc <- parse_factor(c("A", "B"), levels = c("A", "B", "C"))
    
  fct_drop(fc)
    
    ```

- `fct_expand()`: adds a new level.

    ```{r}
     fc <- parse_factor(c("A", "B"))
    
    fct_expand(fc, "C")
    #adds level without changing the vector itself
    ```

- `fct_infreq()`: Order by frequency of a level.

    ```{r}
    fc <- parse_factor(c("A", "B", "C", "A", "B", "C", "C", "B", "B"))
    
    fct_count(fc)
    fct_infreq(fc) %>% 
      fct_count(fc)
    ```




  

